---
output: 
  pdf_document:
    latex_engine: xelatex
---

```{=latex}
\begin{titlepage}
    \centering
    {\Huge\bfseries STA302 Final Project Part 1: Project Proposal \par}
    {\LARGE [Research Question] \par}
    \vspace{0.3cm}
    {\Large Yolanda Thant, Kosar Hemmati, Shameiza Hussain, Ayushi Verma\par}
    {\Large 15 May 2025\par}
\end{titlepage}
```

```{r setup, include=FALSE}
## Hides both code & output in knitted document 
library(tidyverse)  ## For data wrangling & plotting
library(patchwork)
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)  ## Globally suppress all code, messages, & warnings
```

```{r}
## load diamonds dataset
diamonds <- read_csv("./dataset/original_diamonds_dataset.csv")
## list.files("./dataset") 
## getwd()
```

```{r}
## Data Cleaning: Remove predictor variables we don't need: 1st col, cut, depth, y, z
diamonds_cleaned <- diamonds %>% select(-...1, -cut, -depth, -y, -z)
## head(diamonds_cleaned)
write_csv(diamonds_cleaned, "./dataset/diamonds_cleaned.csv")
## summary(diamonds_cleaned)
```

\section*{Introduction}
When considering a diamond’s value, many consumers rely on a grading system introduced by the Gemological Institute of America (n.d.) called the “4Cs”: carat, cut, clarity, and colour. This system was designed to help consumers of the diamond market understand the value of their purchases based on the values assigned to each of these four traits. While information about factors that influence diamond pricing is publicly available, the numeric impact of each factor is unclear (Lee et al., 2014; Mamonov & Triantoro, 2018). For instance, two diamonds of the same weight can differ vastly in price after accounting for cut, clarity, and colour (Tiffany & Co., n.d.), but the explanation behind this variance has not been rigorously quantified based on our literature research. Thus, our research addresses the question: To what extent do carat, colour, clarity, table size, and length predict the price of a diamond? While previous studies have conducted similar analyses (Mamonov & Triantoro, 2018; Özmen, 2024), our goal is to later refine our model in hopes of explaining more of the observed price variation in our data. Quantifying the effects that different features have on diamond prices can better inform consumers’ judgments on whether they are paying a fair price. 

Multiple linear regression (MLR) is a suitable method to answer our question, because we are exploring the association between price, a continuous response variable, with several explanatory variables simultaneously. This way, we can infer which diamond factors are the strongest predictors of price. MLR models the (estimated) average of the response versus given values of the predictors, which is relevant to our goal of helping consumers understand where the price of their diamond lies in relation to the estimated average price for their specific diamond profile. The focus of the model will be on interpretability since we are more interested in drawing conclusions about the average effects of our chosen diamond attributes on the price, rather than trying to predict exact prices of all diamonds. 

\section*{Data Description}
summarize numerically or graphically (in a single figure/table) each predictor in your dataset that
will be used in the preliminary model, and interpret the descrip8ve sta8s8cs in the context of what the predictors
measure and how it relates to the research ques8on

```{r}
## Preliminary (MLR) Model Setup
fit <- lm(price ~ carat + color + clarity + table + x, data=diamonds_cleaned)
## summary(fit)
```

```{r}
## 1(a) Residual vs Fitted Values Plot | NOTE: Best plot among the 3 1(a)(b)(c) to check for 1st 2 assumptions
resid_fit <- data.frame(Fitted = fitted(fit), Residuals = resid(fit))
resid_fitted <- ggplot(data=resid_fit, aes(x=Fitted, y=Residuals)) +
  geom_point(alpha=0.4) +  ## Transparency level to help w/ overlapping pts
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  labs(x="Fitted Values", y="Residuals",
       title="Residuals vs Fitted Values")
```

```{r}
## 1(b) Response vs Fitted Values Plot
price_fitted <- ggplot(data=diamonds_cleaned, aes(x=fitted(fit), y=price)) +
  geom_point(alpha=0.3) +  ## Transparency level to help w/ overlapping pts
  geom_abline(color="blue") +
  labs(x="Fitted Values", y="Actual Price",
       title="Price vs Fitted Values")
```

```{r}
## 1(b) Response vs Predictor Values Plot

## predictors <- c("carat", "color", "clarity", "table", "x")

## price_predictors <- 
##   for (x_i in predictors) {
##   print(
##     ggplot(data=diamonds_cleaned, aes(x=!!sym(x_i), y=price)) +
##       geom_point(alpha=0.2) +
##       geom_smooth(method="lm", color="blue", se = FALSE) +
##       labs(title=paste("Price vs", x_i))
##   )
## }
## !!sym(x_i): Converts str var name into a symbol & unquotes it for aes() to use it directly.

price_carat <- ggplot(data=diamonds_cleaned, aes(x=carat, y=price)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm", color="blue", se=FALSE) +
  labs(title="Price vs Carat")

price_color <- ggplot(data=diamonds_cleaned, aes(x=color, y=price)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm", color="blue", se=FALSE) +
  labs(title="Price vs Color")

## Boxplot helps confirm that color is not constant & shows some variation in price.
## Can use as predictor in modeling price.
price_color_boxplot <- ggplot(data=diamonds_cleaned, aes(x=color, y=price)) +
  geom_boxplot(fill="lightblue", color="black") +
  labs(title="Boxplot of Price by Color",
       x="Color", y="Price")

price_clarity <- ggplot(data=diamonds_cleaned, aes(x=clarity, y=price)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm", color="blue", se=FALSE) +
  labs(title="Price vs Clarity")

## Boxplot confirms that clarity does is not constant & shows meaningful variation in price.
## Can use as predictor in modeling price.
price_clarity_boxplot <- ggplot(data=diamonds_cleaned, aes(x=clarity, y=price)) +
  geom_boxplot(fill="lightblue", color="black") +
  labs(title="Boxplot of Price by Clarity",
       x="Clarity", y="Price")

price_table <- ggplot(data=diamonds_cleaned, aes(x=table, y=price)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm", color="blue", se=FALSE) +
  labs(title="Price vs Table")

price_x <- ggplot(data=diamonds_cleaned, aes(x=x, y=price)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm", color="blue", se=FALSE) +
  labs(title="Price vs x")
```

```{r}
## 2(a) Histogram of Standardized Residuals
std_resid <- rstandard(fit)
sresid_hist <- ggplot(data=data.frame(std_resid), aes(x=std_resid)) +
  geom_histogram(bins=30, fill="lightblue", color="black") +
  labs(title="Histogram of Standardized Residuals",
       x="Standardized Residuals", y="Frequency")
```

```{r}
## Normal QQ-Plot of Standardized Residuals
sresid_qq <- ggplot(data=data.frame(std_resid), aes(sample=std_resid)) +
  stat_qq(alpha=0.2) +
  stat_qq_line(color="red", linetype="dashed") +
  labs(title="Normal QQ-Plot of Standardized Residuals", 
       x="Theoretical Quantiles", y="Standardized Residuals")
```

\section*{Preliminary Model Results}
```{r, fig.width=15, fig.height=20}
(resid_fitted | price_fitted | price_carat) / 
  (price_color | price_color_boxplot) / 
  (price_clarity | price_clarity_boxplot) / 
  (price_table | price_x) / 
  (sresid_hist | sresid_qq)
```

\begin{align*}
y =\ & \beta_0 + \beta_1 \cdot \text{carat} + \beta_2 \cdot I(\text{color} = E) + \beta_3 \cdot I(\text{color} = F) + \beta_4 \cdot I(\text{color} = G) + \beta_5 \cdot I(\text{color} = H) \\
& + \beta_6 \cdot I(\text{color} = I) + \beta_7 \cdot I(\text{color} = J) + \beta_8 \cdot I(\text{clarity} = \text{IF}) + \beta_9 \cdot I(\text{clarity} = \text{SI1}) \\
& + \beta_{10} \cdot I(\text{clarity} = \text{SI2}) + \beta_{11} \cdot I(\text{clarity} = \text{VS1}) + \beta_{12} \cdot I(\text{clarity} = \text{VS2}) \\
& + \beta_{13} \cdot I(\text{clarity} = \text{VVS1}) + \beta_{14} \cdot I(\text{clarity} = \text{VVS2}) + \beta_{15} \cdot \text{table} + \beta_{16} \cdot x + \epsilon \\[0.5em]
\text{} & \text{Assume } \mathbb{E}[\epsilon] = 0, \mathbb{V}[\epsilon] = \sigma^2, \text{and } \epsilon \overset{\text{i.i.d.}}{\sim} N(0, \sigma^2) \\[0.5em]
\hat{E}[y] =\ & \hat{y} =\ \hat{\beta}_0 + \hat{\beta}_1 \cdot \text{carat} + \hat{\beta}_2 \cdot I(\text{color} = E) + \hat{\beta}_3 \cdot I(\text{color} = F) + \hat{\beta}_4 \cdot I(\text{color} = G) + \hat{\beta}_5 \cdot I(\text{color} = H) \\
& + \hat{\beta}_6 \cdot I(\text{color} = I) + \hat{\beta}_7 \cdot I(\text{color} = J) + \hat{\beta}_8 \cdot I(\text{clarity} = \text{IF}) + \hat{\beta}_9 \cdot I(\text{clarity} = \text{SI1}) \\
& + \hat{\beta}_{10} \cdot I(\text{clarity} = \text{SI2}) + \hat{\beta}_{11} \cdot I(\text{clarity} = \text{VS1}) + \hat{\beta}_{12} \cdot I(\text{clarity} = \text{VS2}) \\
& + \hat{\beta}_{13} \cdot I(\text{clarity} = \text{VVS1}) + \hat{\beta}_{14} \cdot I(\text{clarity} = \text{VVS2}) + \hat{\beta}_{15} \cdot \text{table} + \hat{\beta}_{16} \cdot x
\end{align*}

```{r}
## Numerical Summary of Coefficient Estimates
coef_table <- data.frame(
  Term = c("(Intercept)", "carat", "colorE", "colorF", "colorG", "colorH", "colorI", "colorJ",
           "clarityIF", "claritySI1", "claritySI2", "clarityVS1", "clarityVS2", "clarityVVS1", "clarityVVS2",
           "table", "x"),
  Estimate = c(-1388.839, 10945.695, -210.489, -286.752, -493.858, -993.809, -1474.570, -2387.941,
               5665.892, 3894.266, 2926.092, 4845.059, 4528.148, 5309.983, 5234.228, -32.708, -900.270)
)
kable(coef_table, digits=3, caption="Numerical Summary of Coefficient Estimates")

## summary(fit)
```

\section*{Residual Analysis}
The residual analysis revealed significant violations of core regression model assumptions: linearity, constant error variance (homoscedasticity), and normality of errors. The **Residuals vs Fitted Values** plot showed a clear, curved LOESS trendline with increasing spread as fitted values increased, suggesting non-linearity and heteroscedasticity. 

Similarly, the **Price vs Fitted Values** plot showed a generally strong positive relationship, but with slight curvature and growing variance at higher price levels, indicates non-linearity & heteroscedasticity. The **Price vs Carat** scatterplot suggested an exponential relationship that a smile linear model fails to capture reasonably.  

At first glance, categorical variables **color** and **clarity** don’t appear to be strong predictors of price due to seemingly constant spread of points across levels. However, the boxplots reveal significant stratified effects on **price**, confirming that these variables do show variation.

For **color**, although the maximum prices are similar across all levels, the median prices differ, with diamonds in **color** J having higher median prices, and D showing higher outliers.

Diamonds with **clarity** I1 show outliers and higher median prices, whereas those with IF appear to have more outliers and lower medians. This stratification reinforces the idea that both are valid categorical predictors for diamond price modeling.

However, the **Price vs Table** plot showed a weak and noisy association, with data points highly concentrated in a narrow range, suggesting that the table predictor may not be a strong predictor to predict price. The **Price vs x** plot also showed curvature, suggesting non-linearity.

Furthermore, the **histogram of standardized residuals** sharply peaked around 0 with heavy tails, and the **normal QQ-plot** confirmed it as it showed deviations at both tails, indicating violations of the normality assumption. 

\section*{Model Discussion}
From model estimates, a diamond with **color** D, **clarity** I1, and 0 values for **table**, **x**, and **carat**, is predicted to have an average price of -\$1,388.389. Though this scenario isn’t practically meaningful, it serves as the model’s (baseline) reference level. 
When the diamond’s **color** is E and **clarity** is IF, with the same 0 values for the continuous variables, the (estimated) average price increases to \$4,066.564, reflecting the strong positive contribution of high **clarity** IF to the overall **price**.

Moreover, a 1-gram increase in **carat**—with the same values as the previous example—raises the predicted average price by \$10,945.695 per gram of **carat**, confirming that **carat** is the most influential predictor.

Interestingly, **table** and **x** are both negatively associated with **price**, with coefficients of −\$32.71 and −\$900.27, respectively. This is counterintuitive as larger dimensions are generally more desirable. **x** likely reflects multicollinearity with **carat**, since both capture size-related information. This redundancy suggests the need to reconsider variable inclusion—possibility removing **x** or replacing with a composite variable like volume.

```{=latex}
\section*{Bibliography}
\begingroup
\setlength{\parindent}{0pt}
\setlength{\parskip}{8pt}

\hangindent=0.5in Gemological Institute of America. (n.d.). \textit{4Cs of diamond quality}. Retrieved May 20, 2025, from \url{https://4cs.gia.edu/en-us/4cs-of-diamond-quality/}

\hangindent=0.5in Lee, J. A., Caudill, S. B., \& Mixon, F. G. (2014). Shine bright like a diamond: A hedonic model of grading and pricing an experience good. \textit{Applied Economics, 46}(16), 1829--1838. \url{https://doi.org/10.1080/00036846.2014.884707}

\hangindent=0.5in Mamonov, S., \& Triantoro, T. (2018). Subjectivity of diamond prices in online retail: Insights from a data mining study. \textit{Journal of Theoretical and Applied Electronic Commerce Research, 13}(2), 15--28. \url{https://doi.org/10.4067/s0718-18762018000200103}

\hangindent=0.5in \"Ozmen, G. (2024, August 12). \textit{More than carat: A comprehensive regression analysis for diamond price prediction}. Medium. \url{https://medium.com/@gizemzmen/more-than-carat-a-comprehensive-regression-analysis-for-diamond-price-prediction-b9222f337154}

\hangindent=0.5in Tiffany \& Co. (n.d.). \textit{What Is A Diamond Carat?} Retrieved May 20, 2025, from \url{https://www.tiffany.ca/engagement/the-tiffany-guide-to-diamonds/carat/}

\endgroup
```
